# Making Sense of High-Dimensional Data: PCA, t-SNE, and UMAP {#sec-dimensionality-reduction}

```{r}
#| include: false
library(tidyverse)
library(knitr)
```

The previous chapters covered methods that predict an outcome from labelled data through supervised learning. In this chapter, we turn to **unsupervised** methods that explore and summarise data without a specific outcome in mind.

Although many clinical tasks have a clear outcome (e.g., survival, disease progression, treatment response), there are also many situations where the goal is to *explore* a dataset and discover patterns that were not anticipated. For example, you might want to:

- Visualise complex datasets in a way that humans can interpret.
- Identify subgroups of patients with similar disease profiles.
- Detect outliers or unusual cases that merit further investigation.

Such tasks typically use *unsupervised learning* methods, as there is no outcome variable to optimise against. Instead, they optimise a certain metric (e.g., variance explained, local neighbourhood preservation) to find structure in the data.

Unsupervised learning techniques can be generally grouped into three categories depending on the outcome they generate:
- **Dimensionality reduction** methods aim to represent data in a **continuous** low-dimensional space (e.g., PCA, t-SNE, UMAP).
- **Clustering** methods aim to represent data in **discrete** clusters based on similarity (e.g., k-means, hierarchical clustering).
- **Anomaly detection** methods aim to identify **outliers** or unusual observations that do not fit the general pattern of the data.

This course will focus on the first two, which will give you the tools to explore high-dimensional clinical datasets.

## The Curse of Dimensionality

A single patient encounter can generate hundreds of variables: vital signs, lab results, imaging features, medication lists, genomic markers, clinical notes. The (*a priori*) reason for collecting so many variables is that each one may carry some information about the patient's health. However, study cohorts are rarely large enough to support the hundreds, if not thousands, of variables that are measured. In other words, the number of features ($P$) is often large relative to the number of observations ($N$).

This $P \gg N$ scenario is common in modern biomedical research, and can make analysis challenging. Imagine regressing a binary outcome (e.g., disease vs. no disease) on 5,000 gene expression values with only 50 patients. As the model has far more parameters than data points, it risks overfitting the training data and producing unstable estimates. You also risk encountering collinear coefficients that make prediction unstable. The number of false positives increases with each additional predictor. If regression coefficients are unstable, imagine simply visualizing trends in the dataset: how to make sense of 5,000 axes?

Such challenges are collectively referred to as the **curse of dimensionality**: above a certain number of dimensions, our ability to make sense of the data breaks down. Distances between points become less meaningful, and the volume of the space grows so quickly that data points become sparse. This sparsity makes it difficult to find meaningful patterns, as most points are far from each other in high-dimensional space.

To address this issue, **dimensionality reduction** techniques are employed to project high-dimensional data into a lower-dimensional space while preserving as much meaningful structure as possible. This chapter covers three complementary methods: **PCA** (principal component analysis), **t-SNE** (t-distributed stochastic neighbour embedding), and **UMAP** (uniform manifold approximation and projection). Each serves a different purpose, and understanding when to use which is a core skill in modern data analysis.

```{mermaid}
%%| echo: false
%%| eval: true
%%| label: fig-dimreduction-idea
%%| fig-cap: "Motivation for dimensionality reduction: patients in a cohort start as a long list of measurements that cannot be plotted directly. Dimensionality reduction compresses that list into two summary axes, turning every patient into a single dot on a map where similar patients sit close together."
flowchart LR
    A["<b>One patient</b><br/>5,000 measurements<br/>(impossible to plot)"] --> B["<b>Dimensionality<br/>reduction</b><br/>PCA · t-SNE · UMAP"]
    B --> C["<b>One dot</b><br/>on a 2-D map"]
    C --> D["<b>Whole cohort</b><br/>similar patients<br/>cluster together"]
    style A fill:#eef3fb,stroke:#4a6fa5
    style B fill:#fbf3e6,stroke:#c79a3b
    style C fill:#eef3fb,stroke:#4a6fa5
    style D fill:#e8f1ea,stroke:#4a7a55
```

## PCA: Principal Component Analysis

PCA is the oldest and most widely used dimensionality reduction technique. It aims to find directions (also called *principal components*) that maximise the variance in this new coordinate system. It has two important features:
- Each principal component is a linear combination of the original variables. For example, for a metabolic panel, PC1 might be a weighted sum of glucose, HbA1c, and triglycerides.
- Each principal component is orthogonal to the others, meaning they capture independent sources of variation in the data.

### The Core Idea

We start with a data matrix $\mathbf{X}$ with $N$ observations and $P$ variables, and assume that the data come from i.i.d. (independent and identically distributed) samples. To find the principal components, we can follow these steps:

1. **Centre** the data: subtract the mean of each variable. Experience also shows that it is often useful to also divide by the standard deviation of each variable (standardisation), especially when variables are on different scales.
We obtain $Z$ such that:
$$Z_{ij} = \frac{X_{ij} - \bar{X}_j}{\sigma_j}$$
where $\bar{X}_j$ is the mean of variable $j$ and $\sigma_j$ is its standard deviation.
2. **Compute the covariance matrix** $\mathbf{C} = \frac{1}{N}\mathbf{Z}^T\mathbf{Z}$.
3. **Find the eigendecomposition** of $\mathbf{C}$. The eigenvectors are the **principal components** (the directions of maximum variance), the eigenvalues are the **amount of variance explained** by each component.

By construction, the first principal component (PC1) captures the most variance, PC2 captures the second most (orthogonal to PC1), and so on. By keeping only the first $K$ components, you reduce the data from $P$ dimensions to $K$ dimensions.

In plain terms, a **principal component** is just a recipe: a new variable made by mixing the original measurements in fixed proportions (the weights in that recipe are called the **loadings**). PC1 is the single mixture that spreads patients out as much as possible (the direction along which they differ most). PC2 is the direction, perpendicular to PC1, that captures the most remaining spread. Each subsequent component captures less and less, because the "best" directions are always taken first.

Going back to our example, if glucose, HbA1c, and triglycerides all rise and fall together, PCA notices and bundles them into one axis, so a patient's position on PC1 might summarise their whole "metabolic" profile in a single number. "Variance explained" is then simply the share of the data's total spread that each component accounts for: PC1 might capture 40% of all the variation, meaning that one axis alone tells you a large chunk of what distinguishes your patients.

### Key Properties

- PCA is a **linear** method: each component is a weighted sum of the original variables.
- It is **deterministic**: running PCA twice on the same data gives the same result (up to sign flips).
- It preserves **global structure** (distances and variance) well, but may miss non-linear patterns (as it is a linear method).

### Choosing the Number of Components

Choosing the number of components depends on the context and the goal of the analysis. Common strategies include:

- **Scree plot.** Plot the eigenvalues (or proportion of variance explained) against the component number. Look for an "elbow," a point where the curve flattens, suggesting that additional components contribute little.

- **Cumulative variance threshold.** Keep enough components to explain a target percentage of total variance (commonly 90--95--99%).

### Clinical Example: Metabolic Panel PCA

A common application: reducing a panel of correlated lab values to a few interpretable dimensions. Consider a dataset with glucose, HbA1c, triglycerides, LDL, HDL, creatinine, ALT, and albumin.

::: {.panel-tabset}

### R

```{r}
#| label: fig-pca-metabolic-r
#| fig-cap: "PCA of a simulated metabolic panel: scree plot and biplot."
#| fig-width: 12
#| fig-height: 5

set.seed(42)
n <- 300

# Simulate correlated metabolic data
# Create correlation structure: glucose/HbA1c/triglycerides cluster;
# creatinine/albumin cluster
library(MASS)
# fmt: skip
Sigma <- matrix(c(
  1.0, 0.7, 0.5, 0.3,-0.2, 0.1, 0.2,-0.1,
  0.7, 1.0, 0.4, 0.2,-0.2, 0.1, 0.1,-0.1,
  0.5, 0.4, 1.0, 0.4,-0.3, 0.1, 0.3,-0.1,
  0.3, 0.2, 0.4, 1.0,-0.5, 0.1, 0.2, 0.0,
 -0.2,-0.2,-0.3,-0.5, 1.0,-0.1,-0.1, 0.2,
  0.1, 0.1, 0.1, 0.1,-0.1, 1.0, 0.1,-0.3,
  0.2, 0.1, 0.3, 0.2,-0.1, 0.1, 1.0,-0.2,
 -0.1,-0.1,-0.1, 0.0, 0.2,-0.3,-0.2, 1.0
), nrow = 8)

z <- mvrnorm(n, mu = rep(0, 8), Sigma = Sigma)
metabolic <- tibble(
  glucose = round(z[, 1] * 30 + 100),
  hba1c = round(z[, 2] * 1.0 + 5.8, 1),
  triglycerides = round(z[, 3] * 50 + 150),
  ldl = round(z[, 4] * 30 + 120),
  hdl = round(z[, 5] * 12 + 55),
  creatinine = round(z[, 6] * 0.3 + 1.0, 2),
  alt = round(z[, 7] * 15 + 30),
  albumin = round(z[, 8] * 0.4 + 4.0, 1)
)

# PCA on scaled data
pca_result <- prcomp(metabolic, scale. = TRUE)

# Scree plot and biplot side by side
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Scree plot
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
barplot(
  var_explained,
  names.arg = paste0("PC", 1:8),
  col = "steelblue",
  main = "Scree Plot",
  ylab = "Proportion of Variance",
  xlab = "Component",
  ylim = c(0, 0.4)
)
# Biplot
biplot(
  pca_result,
  scale = 0,
  cex = 0.6,
  col = c("grey70", "firebrick"),
  main = "PCA Biplot: Metabolic Panel"
)
```

### Python

```{python}
#| label: fig-pca-metabolic-py
#| fig-cap: "PCA of a simulated metabolic panel: scree plot and biplot."

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA

np.random.seed(42)
n = 300

Sigma = np.array(
    [
        [1.0, 0.7, 0.5, 0.3, -0.2, 0.1, 0.2, -0.1],
        [0.7, 1.0, 0.4, 0.2, -0.2, 0.1, 0.1, -0.1],
        [0.5, 0.4, 1.0, 0.4, -0.3, 0.1, 0.3, -0.1],
        [0.3, 0.2, 0.4, 1.0, -0.5, 0.1, 0.2, 0.0],
        [-0.2, -0.2, -0.3, -0.5, 1.0, -0.1, -0.1, 0.2],
        [0.1, 0.1, 0.1, 0.1, -0.1, 1.0, 0.1, -0.3],
        [0.2, 0.1, 0.3, 0.2, -0.1, 0.1, 1.0, -0.2],
        [-0.1, -0.1, -0.1, 0.0, 0.2, -0.3, -0.2, 1.0],
    ]
)

z = np.random.multivariate_normal(np.zeros(8), Sigma, n)
labels = [
    "glucose",
    "hba1c",
    "triglycerides",
    "ldl",
    "hdl",
    "creatinine",
    "alt",
    "albumin",
]
metabolic = pd.DataFrame(z, columns=labels)

# Scale and fit PCA
scaler = StandardScaler()
X_scaled = scaler.fit_transform(metabolic)
pca = PCA()
scores = pca.fit_transform(X_scaled)

fig, axes = plt.subplots(1, 2, figsize=(14, 5))

# Scree plot
axes[0].bar(
    range(1, 9), pca.explained_variance_ratio_, color="steelblue", edgecolor="white"
)
axes[0].set_xlabel("Component")
axes[0].set_ylabel("Proportion of Variance")
axes[0].set_title("Scree Plot")
axes[0].set_xticks(range(1, 9))

# Biplot
loadings = pca.components_[:2].T  # first 2 PCs
scale_factor = 3
axes[1].scatter(scores[:, 0], scores[:, 1], alpha=0.3, s=10, color="grey")
for i, lab in enumerate(labels):
    axes[1].annotate(
        "",
        xy=(loadings[i, 0] * scale_factor, loadings[i, 1] * scale_factor),
        xytext=(0, 0),
        arrowprops=dict(arrowstyle="->", color="firebrick", lw=1.5),
    )
    axes[1].text(
        loadings[i, 0] * scale_factor * 1.15,
        loadings[i, 1] * scale_factor * 1.15,
        lab,
        fontsize=9,
        color="firebrick",
        ha="center",
    )
axes[1].set_xlabel(f"PC1 ({pca.explained_variance_ratio_[0]:.1%} var)")
axes[1].set_ylabel(f"PC2 ({pca.explained_variance_ratio_[1]:.1%} var)")
axes[1].set_title("PCA Biplot: Metabolic Panel")
axes[1].axhline(0, color="grey", linewidth=0.5)
axes[1].axvline(0, color="grey", linewidth=0.5)

plt.tight_layout()
plt.show()
```

:::

**What the code is doing.** We simulate 300 patients with eight correlated lab values, standardise them (so glucose in the hundreds does not drown out HbA1c around 6), and run PCA. The code then draws two figures. The **scree plot** (left) is a bar chart of how much variance each of the eight components captures, tallest on the left;  You would look for the bars to drop off sharply after the first two or three, telling you most of the signal lives in just a few dimensions. The **biplot** (right) places every patient on the PC1--PC2 plane and overlays arrows for the original labs; how to read those arrows is explained next.

### Interpreting the Biplot

The biplot overlays two things:

- **Points** (grey): individual patients projected onto the first two principal components.
- **Arrows** (red): the original variables. The direction and length of each arrow show how strongly and in which direction that variable contributes to each PC.

In our metabolic panel:

- **PC1** is dominated by glucose, HbA1c, and triglycerides pointing in one direction, with HDL pointing in the opposite direction. This is a "metabolic syndrome" axis: patients to the right tend to have higher glucose and triglycerides and lower HDL.
- **PC2** separates creatinine and albumin, capturing a "renal/hepatic" dimension.

This kind of interpretation is exactly what makes PCA useful in clinical research: it reveals which variables "travel together" and suggests underlying biological dimensions.

## t-SNE: Non-Linear Dimensionality Reduction

### The Core Idea

PCA preserves global, linear structure. But many datasets contain non-linear patterns that PCA cannot capture well. **t-SNE** ($t$-distributed stochastic neighbour embedding) was designed specifically for **visualisation** of high-dimensional data in 2 or 3 dimensions.

t-SNE works in two steps:

1. In the high-dimensional space, compute the probability that any two points are "neighbours" (using a Gaussian kernel). Nearby points have high probability; distant points have low probability.
2. In the low-dimensional (2D) embedding, try to arrange points so that the same neighbourhood probabilities are preserved (using a **t-distribution** kernel, which has heavier tails than a Gaussian and thus helps prevent crowding).

Intuitively, t-SNE asks "who are each patient's closest neighbours?" in the full high-dimensional space, then arranges the 2D plot so that those same neighbourly relationships are preserved. The payoff for a health researcher is a map where patients who are genuinely alike clump together, making subtypes pop out visually. This has made t-SNE become popular for single-cell and other omics data where distinct cell populations or patient phenotypes hide in hundreds of dimensions.

### The Perplexity Parameter

The most important parameter in t-SNE is **perplexity** (typically 5--50), which controls how many neighbours each point considers. It roughly corresponds to the effective number of local neighbours.

- **Low perplexity** (5--10): focuses on very local structure. May produce many small, tight clusters.
- **High perplexity** (30--50): considers broader neighbourhoods. Produces smoother, more global layouts.

There is no single correct value; try several and see how the structure changes.

### Critical Pitfalls and Misinterpretations

::: {.callout-warning}
## t-SNE is for exploration, not inference
t-SNE is a visualisation tool. It is not a statistical test. Do not draw conclusions about statistical significance from a t-SNE plot.
:::

**Distances between clusters are meaningless.** Two clusters that appear far apart may not actually be more different than two clusters that appear close. t-SNE distorts global distances to preserve local neighbourhoods.

**Cluster sizes are meaningless.** A tight cluster in t-SNE space does not mean those points are more similar than a spread-out cluster.

**Different runs give different layouts.** t-SNE uses stochastic optimisation. Running it twice with different seeds gives different-looking plots (though the same clusters should emerge).

**It does not scale well.** Standard t-SNE compares every patient with every other patient, so the amount of computation grows *quadratically* with the sample size. In practice it makes plain t-SNE notably slow beyond ~10,000 patients.

### Clinical Example: Visualising Patient Phenotypes

::: {.panel-tabset}

### R

```{r}
#| label: fig-tsne-r
#| fig-cap: "t-SNE visualisation of simulated patient data with three disease subtypes."
#| fig-width: 10
#| fig-height: 5

library(Rtsne)

set.seed(42)

# Simulate 3 patient subtypes with 20 features
n_per_group <- 100
p <- 20

group1 <- matrix(rnorm(n_per_group * p, mean = 0, sd = 1), ncol = p)
group2 <- matrix(rnorm(n_per_group * p, mean = 2, sd = 1.2), ncol = p)
group3 <- matrix(rnorm(n_per_group * p, mean = -1, sd = 0.8), ncol = p)
# Make groups differ in specific feature subsets
group2[, 1:5] <- group2[, 1:5] + 3
group3[, 10:15] <- group3[, 10:15] - 4

X <- rbind(group1, group2, group3)
labels <- factor(rep(c("Type A", "Type B", "Type C"), each = n_per_group))

# Run t-SNE with two perplexity values
tsne_low <- Rtsne(
  X,
  perplexity = 10,
  dims = 2,
  verbose = FALSE,
  max_iter = 1000
)
tsne_high <- Rtsne(
  X,
  perplexity = 40,
  dims = 2,
  verbose = FALSE,
  max_iter = 1000
)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
cols <- c("steelblue", "firebrick", "forestgreen")

plot(
  tsne_low$Y,
  col = cols[labels],
  pch = 16,
  cex = 0.8,
  main = "t-SNE (perplexity = 10)",
  xlab = "t-SNE 1",
  ylab = "t-SNE 2"
)
legend("topright", levels(labels), col = cols, pch = 16, cex = 0.8)

plot(
  tsne_high$Y,
  col = cols[labels],
  pch = 16,
  cex = 0.8,
  main = "t-SNE (perplexity = 40)",
  xlab = "t-SNE 1",
  ylab = "t-SNE 2"
)
legend("topright", levels(labels), col = cols, pch = 16, cex = 0.8)
```

### Python

```{python}
#| label: fig-tsne-py
#| fig-cap: "t-SNE visualisation of simulated patient data with three disease subtypes."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.manifold import TSNE

np.random.seed(42)
n_per = 100
p = 20

g1 = np.random.normal(0, 1, (n_per, p))
g2 = np.random.normal(2, 1.2, (n_per, p))
g3 = np.random.normal(-1, 0.8, (n_per, p))
g2[:, :5] += 3
g3[:, 10:15] -= 4

X = np.vstack([g1, g2, g3])
labels = np.repeat(["Type A", "Type B", "Type C"], n_per)
colors = {"Type A": "steelblue", "Type B": "firebrick", "Type C": "forestgreen"}

fig, axes = plt.subplots(1, 2, figsize=(12, 5))
for ax, perp in zip(axes, [10, 40]):
    tsne = TSNE(n_components=2, perplexity=perp, random_state=42, max_iter=1000)
    emb = tsne.fit_transform(X)
    for lab in ["Type A", "Type B", "Type C"]:
        mask = labels == lab
        ax.scatter(emb[mask, 0], emb[mask, 1], c=colors[lab],
                   s=15, alpha=0.7, label=lab)
    ax.set_title(f"t-SNE (perplexity = {perp})")
    ax.set_xlabel("t-SNE 1")
    ax.set_ylabel("t-SNE 2")
    ax.legend(fontsize=9)

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** We build a toy dataset of 300 patients across three "disease subtypes" (each subtype differs on a different handful of the 20 features), then run t-SNE twice: once with perplexity 10, once with 40. We then plot the two resulting maps side by side, colouring each dot by its true subtype. Each dot is a patient; the axes ("t-SNE 1/2") have no units and no inherent meaning, so do not try to read values off them. What you look for is *separation*: three coloured blobs that do not overlap means the subtypes really do have distinct profiles.

Notice how perplexity changes the visual appearance but the three groups are consistently separated. This is reassuring: the cluster structure is real, not an artefact of a particular perplexity setting. (The perplexity setting, recall, is roughly the number of near neighbours each point pays attention to: a small value zooms in on tight local clumps, a large value smooths things into broader groupings.)

## UMAP: Uniform Manifold Approximation and Projection

### How UMAP Differs from t-SNE

UMAP (McInnes, Healy, and Melville, 2018) is a newer method that has can be used as an alternative to t-SNE in many applications. A key advantage is its speed, as it runs 3--10x faster than t-SNE, and its better stability across runs.  UMAP works by constructing a topological representation (a weighted graph) of the data in high dimensions, then optimising a low-dimensional layout that preserves the graph structure. While it has been claimed that UMAP preserves more global structure than t-SNE, this has been disputed in recent literature [@chari2023].


### Key Parameters

**n_neighbors** (analogous to perplexity in t-SNE): controls the balance between local and global structure. Small values (5--15) emphasise local clusters; large values (50--200) emphasise global connectivity.

**min_dist**: controls how tightly points are packed in the embedding. Small values (0.0--0.1) produce tight, well-separated clusters; large values (0.5--1.0) produce a more uniform spread.

### When to Use Which Method

- **PCA**: use for initial exploration, as a preprocessing step (reduce to 20--50 PCs before t-SNE/UMAP), or when you need interpretable components. Use when linear relationships are sufficient.
- **t-SNE and UMAP**: use for detailed visualisation of moderate-sized datasets (< 10,000 points) when you want to emphasise local cluster structure. Avoid for quantitative downstream analysis.

::: {.panel-tabset}

### R

```{r}
#| label: fig-umap-comparison-r
#| fig-cap: "PCA, t-SNE, and UMAP visualisations of the same simulated patient dataset."
#| fig-width: 14
#| fig-height: 4.5

library(Rtsne)
library(uwot)

set.seed(42)

pca_scores <- prcomp(X, scale. = TRUE)$x[, 1:2]
tsne_emb <- Rtsne(X, perplexity = 30, dims = 2, verbose = FALSE)$Y
umap_emb <- umap(X, n_neighbors = 15, min_dist = 0.1, verbose = FALSE)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
cols <- c("steelblue", "firebrick", "forestgreen")[as.numeric(labels)]

plot(
  pca_scores,
  col = cols,
  pch = 16,
  cex = 0.8,
  main = "PCA",
  xlab = "PC1",
  ylab = "PC2"
)

plot(
  tsne_emb,
  col = cols,
  pch = 16,
  cex = 0.8,
  main = "t-SNE (perplexity = 30)",
  xlab = "t-SNE 1",
  ylab = "t-SNE 2"
)

plot(
  umap_emb,
  col = cols,
  pch = 16,
  cex = 0.8,
  main = "UMAP (n_neighbors = 15)",
  xlab = "UMAP 1",
  ylab = "UMAP 2"
)
```

### Python

```{python}
#| label: fig-umap-comparison-py
#| fig-cap: "PCA, t-SNE, and UMAP visualisations of the same simulated patient dataset."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.decomposition import PCA
from sklearn.manifold import TSNE
from umap import UMAP

pca_2d = PCA(n_components=2).fit_transform(X)
tsne_2d = TSNE(n_components=2, perplexity=30, random_state=42).fit_transform(X)
umap_2d = UMAP(n_components=2, n_neighbors=15, min_dist=0.1,
                random_state=42).fit_transform(X)

fig, axes = plt.subplots(1, 3, figsize=(16, 5))
titles = ["PCA", "t-SNE (perplexity=30)", "UMAP (n_neighbors=15)"]
embeddings = [pca_2d, tsne_2d, umap_2d]

for ax, emb, title in zip(axes, embeddings, titles):
    for lab, col in colors.items():
        mask = labels == lab
        ax.scatter(emb[mask, 0], emb[mask, 1], c=col, s=15, alpha=0.7, label=lab)
    ax.set_title(title)
    ax.set_xlabel("Dim 1")
    ax.set_ylabel("Dim 2")
    ax.legend(fontsize=8)

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** We take the *same* three-subtype dataset and run all three methods on it (PCA, t-SNE, and UMAP), then plot their 2D outputs in one row so you can compare them like for like, with patients coloured by true subtype. The point of the figure is not any single panel but the contrast between them: it lets you see, on real-looking data, how a linear method (PCA) and two non-linear methods behave on the same problem.

All three methods recover the three groups, but with different visual characteristics. PCA shows the groups with some overlap (because the separation is partly non-linear). t-SNE and UMAP separate them more clearly, though the visual differences between the two non-linear methods are often modest.

## Practical Workflow: Scale, Reduce, Visualise, Interpret

A standard workflow for high-dimensional clinical data runs in four steps, summarised below.

```{mermaid}
%%| echo: false
%%| eval: true
%%| label: fig-dimreduction-workflow
%%| fig-cap: "The standard four-step pipeline. PCA does the heavy lifting of noise removal and compression; the non-linear method (UMAP or t-SNE) then draws the human-readable map. This PCA-then-UMAP pipeline follows the practice recommended by @mcinnes2018umap."
flowchart LR
    S["<b>1. Scale</b><br/>mean 0, SD 1<br/>so big-unit labs<br/>don't dominate"] --> P["<b>2. Reduce with PCA</b><br/>hundreds of features<br/>→ 20–50 components<br/>(strips noise, speeds things up)"]
    P --> U["<b>3. Embed</b><br/>UMAP or t-SNE<br/>→ 2-D map"]
    U --> I["<b>4. Interpret</b><br/>look for clusters,<br/>gradients, outliers<br/>— then confirm formally"]
    style S fill:#fbf3e6,stroke:#c79a3b
    style P fill:#eef3fb,stroke:#4a6fa5
    style U fill:#eef3fb,stroke:#4a6fa5
    style I fill:#e8f1ea,stroke:#4a7a55
```

1. **Scale** the data. PCA, t-SNE, and UMAP are all sensitive to the scale of variables. Always standardise (mean = 0, SD = 1) before applying any dimensionality reduction.

2. **Reduce with PCA first.** If the original data have hundreds or thousands of features, first reduce to 20--50 principal components. This removes noise and speeds up t-SNE/UMAP dramatically.

3. **Visualise with UMAP (or t-SNE).** Apply the non-linear method to the PCA-reduced data. Colour points by known labels (diagnosis, treatment group) or by continuous variables (lab values, scores).

4. **Interpret cautiously.** Look for clusters, gradients, or outliers. Generate hypotheses, but remember that these are visualisations, not tests. Validate any apparent structure with formal statistical methods.

::: {.panel-tabset}

### R

```{r}
#| label: fig-workflow-r
#| fig-cap: "The standard workflow: PCA for initial reduction, then UMAP for visualisation."
#| fig-width: 10
#| fig-height: 5

library(uwot)

set.seed(42)

# Simulate high-dimensional data: 500 patients, 100 features
n <- 500
p <- 100
group <- sample(
  c("Healthy", "Mild", "Severe"),
  n,
  replace = TRUE,
  prob = c(0.4, 0.35, 0.25)
)

X_hd <- matrix(rnorm(n * p), ncol = p)
# Add signal in first few features
X_hd[group == "Mild", 1:10] <- X_hd[group == "Mild", 1:10] + 1.5
X_hd[group == "Severe", 1:10] <- X_hd[group == "Severe", 1:10] + 3
X_hd[group == "Severe", 11:20] <- X_hd[group == "Severe", 11:20] - 2

# Step 1: Scale
X_scaled <- scale(X_hd)

# Step 2: PCA to 20 components
pca_20 <- prcomp(X_scaled)$x[, 1:20]

# Step 3: UMAP on PCA-reduced data
umap_result <- umap(pca_20, n_neighbors = 15, min_dist = 0.1, verbose = FALSE)

# Step 4: Visualise
plot_df <- tibble(
  UMAP1 = umap_result[, 1],
  UMAP2 = umap_result[, 2],
  Group = factor(group, levels = c("Healthy", "Mild", "Severe"))
)

ggplot(plot_df, aes(x = UMAP1, y = UMAP2, colour = Group)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_colour_manual(
    values = c(
      "Healthy" = "steelblue",
      "Mild" = "goldenrod",
      "Severe" = "firebrick"
    )
  ) +
  labs(
    title = "UMAP After PCA Reduction",
    subtitle = "500 patients, 100 features → 20 PCs → 2D UMAP"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Python

```{python}
#| label: fig-workflow-py
#| fig-cap: "The standard workflow: PCA for initial reduction, then UMAP for visualisation."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA
from umap import UMAP

np.random.seed(42)
n, p = 500, 100
group = np.random.choice(["Healthy", "Mild", "Severe"], n, p=[0.4, 0.35, 0.25])

X_hd = np.random.normal(0, 1, (n, p))
X_hd[group == "Mild", :10] += 1.5
X_hd[group == "Severe", :10] += 3
X_hd[group == "Severe", 10:20] -= 2

# Workflow
X_scaled = StandardScaler().fit_transform(X_hd)
pca_20 = PCA(n_components=20).fit_transform(X_scaled)
umap_2d = UMAP(n_neighbors=15, min_dist=0.1, random_state=42).fit_transform(pca_20)

fig, ax = plt.subplots(figsize=(8, 6))
color_map = {"Healthy": "steelblue", "Mild": "goldenrod", "Severe": "firebrick"}
for g in ["Healthy", "Mild", "Severe"]:
    mask = group == g
    ax.scatter(umap_2d[mask, 0], umap_2d[mask, 1],
               c=color_map[g], s=12, alpha=0.6, label=g)
ax.set_xlabel("UMAP 1")
ax.set_ylabel("UMAP 2")
ax.set_title("UMAP After PCA Reduction\n500 patients, 100 features → 20 PCs → 2D")
ax.legend()
plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** This puts the four-step recipe into practice on a larger, more realistic problem: 500 patients, 100 features, and three severity groups (Healthy, Mild, Severe) whose differences are buried in only the first 20 features. We scale, compress the 100 features down to 20 principal components, then hand those to UMAP to draw a single 2D map, colouring each patient by severity. You would expect to see the three colours form a gradient or three loose clusters (Healthy at one end, Severe at the other), which would confirm that the severity signal survived the squeeze from 100 dimensions to 2. Doing PCA first is not just for speed: it strips out noise so UMAP focuses on the real structure.

## Exercises

::: {.callout-tip title="Exercise 1: PCA on Clinical Lab Data"}

Using the simulated metabolic panel data from this chapter (or a real dataset if available):

a. Perform PCA on the standardised data.
b. Create a scree plot and determine how many components to retain using the cumulative variance threshold (80%).
c. Examine the loadings of the first two PCs. Which variables load most strongly on each? Propose a clinical interpretation.
d. Create a biplot coloured by a simulated diabetes status variable. Do diabetic patients separate along PC1?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}

### R

```{r}
#| label: exercise-1-solution
#| fig-width: 10
#| fig-height: 5

library(tidyverse) # %>% pipe, select()

# Add diabetes status (correlated with glucose/HbA1c)
set.seed(99)
metabolic$diabetes <- factor(ifelse(
  metabolic$glucose > 110 & metabolic$hba1c > 6.2,
  "Diabetic",
  "Non-diabetic"
))

pca_fit <- prcomp(metabolic %>% select(-diabetes), scale. = TRUE)

# (a) & (b) Scree plot with cumulative variance
var_prop <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
cum_var <- cumsum(var_prop)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
barplot(
  var_prop,
  names.arg = paste0("PC", 1:8),
  col = "steelblue",
  main = "Variance Explained",
  ylab = "Proportion"
)
abline(h = 1 / 8, col = "firebrick", lty = 2)

plot(
  1:8,
  cum_var,
  type = "b",
  pch = 16,
  col = "steelblue",
  main = "Cumulative Variance",
  xlab = "Components",
  ylab = "Cumulative proportion",
  ylim = c(0, 1)
)
abline(h = 0.80, col = "firebrick", lty = 2)
text(6, 0.82, "80% threshold", col = "firebrick", cex = 0.8)

cat("Components for 80% variance:", which(cum_var >= 0.80)[1], "\n")

# (c) Loadings: how strongly each variable contributes to PC1 and PC2
loadings_tab <- data.frame(
  Variable = rownames(pca_fit$rotation),
  PC1 = round(pca_fit$rotation[, 1], 3),
  PC2 = round(pca_fit$rotation[, 2], 3),
  row.names = NULL
)
knitr::kable(
  loadings_tab[order(-abs(loadings_tab$PC1)), ],
  row.names = FALSE,
  caption = "Variable loadings on the first two principal components (sorted by absolute PC1 loading)."
)
```

### Python

```{python}
#| label: exercise-1-solution-py

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA

# Recreate metabolic data (using earlier arrays)
diabetes = ["Diabetic" if metabolic.iloc[i]['glucose'] > 110 and
            metabolic.iloc[i]['hba1c'] > 6.2 else "Non-diabetic"
            for i in range(len(metabolic))]

X = metabolic[['glucose','hba1c','triglycerides','ldl','hdl',
               'creatinine','alt','albumin']].values
X_s = StandardScaler().fit_transform(X)
pca = PCA().fit(X_s)

cum_var = np.cumsum(pca.explained_variance_ratio_)
n_80 = np.argmax(cum_var >= 0.80) + 1
print(f"Components for 80% variance: {n_80}")

# Loadings
feat_names = ['glucose','hba1c','triglycerides','ldl','hdl',
              'creatinine','alt','albumin']
for pc in [0, 1]:
    loadings = pca.components_[pc]
    order = np.argsort(np.abs(loadings))[::-1]
    print(f"\nPC{pc+1} loadings:")
    for idx in order:
        print(f"  {feat_names[idx]:>15s}: {loadings[idx]:+.3f}")
```

:::

**What the solution shows.** The code labels patients as diabetic or not (based on glucose and HbA1c), re-runs PCA, and prints two things: how many components are needed to reach 80% of the variance, and the loadings of PC1 and PC2 sorted by size. Reading the printed loadings is the key step: the labs with the largest positive or negative numbers are the ones that define each axis. If glucose, HbA1c, and triglycerides top the PC1 list, then PC1 is effectively a "metabolic" score, and you would expect diabetic patients to sit toward one end of that axis.
:::

:::

::: {.callout-tip title="Exercise 2: t-SNE Sensitivity to Perplexity"}

Using the three-group simulated dataset:

a. Run t-SNE with perplexity values of 5, 15, 30, and 50.
b. Create a 2x2 panel of the resulting embeddings.
c. At which perplexity value do the three groups first become clearly separated?
d. Are the distances between clusters consistent across perplexity values? What does this tell you about interpreting t-SNE?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch15_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch15_ex2.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 3: UMAP Parameter Exploration"}

Using the same dataset:

a. Run UMAP with `n_neighbors` in {5, 15, 50, 100} while holding `min_dist = 0.1`.
b. Run UMAP with `min_dist` in {0.0, 0.1, 0.5, 1.0} while holding `n_neighbors = 15`.
c. Create a panel of plots for each parameter sweep.
d. Describe how each parameter affects the visual appearance. Which combination would you recommend for this dataset?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch15_ex3.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch15_ex3.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 4: Full Workflow on Simulated Genomic Data"}

Simulate a dataset of 1,000 patients with 500 gene expression features and 4 cancer subtypes.

a. Scale the data and perform PCA. How many PCs are needed for 80% variance?
b. Apply UMAP to the first 30 PCs. Colour by cancer subtype. Are the subtypes visually separable?
c. Repeat with t-SNE. Compare the two visualisations.
d. One of the four subtypes is rare (5% of patients). Can you still identify it in the UMAP plot?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch15_ex4.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch15_ex4.py
```
:::
:::

:::

## Summary

Dimensionality reduction transforms high-dimensional clinical data into interpretable, lower-dimensional representations. PCA provides linear, interpretable components that are ideal for initial exploration and preprocessing. t-SNE and UMAP offer non-linear embeddings that reveal complex structure invisible to PCA, though their outputs must be interpreted with care. The standard workflow (scale, reduce with PCA, embed with UMAP or t-SNE, interpret cautiously) is a reliable starting point for any high-dimensional dataset in clinical research.

::: {.callout-tip}
## Key Takeaways
1. Always scale your data before dimensionality reduction.
2. PCA is linear, deterministic, and interpretable. Use it first.
3. t-SNE and UMAP are for visualisation only. Distances between clusters are meaningless; cluster sizes are meaningless.
4. Non-linear methods are exploratory. Validate any apparent structure with formal statistical methods or domain expertise.
:::

## References and Further Reading

- Jolliffe and Cadima, [*Principal Component Analysis: A Review and Recent Developments*](https://doi.org/10.1098/rsta.2015.0202) (Philosophical Transactions of the Royal Society A, 2016). A modern overview of PCA covering both classical theory and recent extensions.
- van der Maaten and Hinton, [*Visualizing Data using t-SNE*](https://jmlr.org/papers/v9/vandermaaten08a.html) (Journal of Machine Learning Research, 2008). The original t-SNE paper.
- Wattenberg, Viegas, and Johnson, [*How to Use t-SNE Effectively*](https://distill.pub/2016/misread-tsne/) (Distill, 2016). An interactive guide to t-SNE's behaviour and pitfalls. Essential reading before interpreting t-SNE plots.
- McInnes, Healy, and Melville, [*UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction*](https://arxiv.org/abs/1802.03426) (arXiv, 2018). The foundational UMAP paper.
- Chari and Pachter, [*The Specious Art of Single-Cell Genomics*](https://doi.org/10.1371/journal.pcbi.1011288) (PLOS Computational Biology, 2023). A critical assessment showing that both t-SNE and UMAP induce substantial distortions; challenges claims that UMAP better preserves global structure.
- Altman and Krzywinski, [*The Curse(s) of Dimensionality*](https://doi.org/10.1038/s41592-018-0019-x) (Nature Methods, 2018). A concise primer on why high-dimensional data are inherently difficult to analyse.
- James, Witten, Hastie, and Tibshirani, [*An Introduction to Statistical Learning*](https://www.statlearning.com/) (2nd edition, Springer, 2021). Chapter 12 provides an accessible introduction to PCA well suited for students with a biostatistics background.

::: {#refs}
:::
